Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-03-26 analysis/paper/paper.qmd
minor edits
finish editing

We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that the tools were being used as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variations in patch productivity and travel times. The variability of TCSA values in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters. After the LGM we anticipate a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.

@@

```{r}
#| label: fig-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A: Modern vegetation map defined by temperature and precipitation. The individual points indicate assemblages. B: TCSA distribution for vegetation zones. The individual points indicate artifacts. C: Tukey test result for vegetation zones. D: Eco-region map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). E: TCSA distribution for ecoregion zones, excluding zones with no artifacts. F: Tukey test result for eco-region zones"
#| fig-height:  5

@@

Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`). This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.

@@

Overall, our results show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. The widest TCSA range was found at a single site, Suyanggae (@fig-tcsa-per-sites). Our findings are therefore consistent with our second scenario, in which stemmed points are best described as multifunctional tools, suggesting that in most cases people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explained Clovis points. The LGM period may be the exception here, with a reduction in variability during this time.

@@

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Palaeolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have frequently encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.

>>>>

We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that the tools were being used as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variations in patch productivity and travel times. The variability of TCSA values, as measured by coefficients of variation (CV), in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters. After the LGM we anticipate a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.

@@

```{r}
#| label: fig-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A: Modern vegetation map defined by temperature and precipitation. The individual points indicate assemblages. B: TCSA distribution for vegetation zones. The individual points indicate artifacts. C: Tukey test result for vegetation zones. D: Eco-region map based on geographical boundaries and ecological conditions (modified from @lee2008relationship). E: TCSA distribution for ecoregion zones, excluding zones with no artifacts. F: Tukey test result for eco-region zones"
#| fig-height:  5

@@

Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Contrary to our prediction, the earliest stemmed points do not have more variable TSCA values than subsequent ones. The most striking temporal pattern is simply that stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`). This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.

@@

Overall, our results show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. The widest TCSA range was found at a single site, Suyanggae (@fig-tcsa-per-sites). Our findings are therefore consistent with our second scenario, in which stemmed points are best described as multifunctional tools, suggesting that in most cases people created stemmed points as a response to unexpected or varying circumstances in their specific habitats, similar to how @eren2022plains explained Clovis points. The LGM period may be the exception here, with a reduction in variability during this time.

@@

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Palaeolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic. Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks. During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged. Considering that many of our predictions about TCSA were not supported by the data, we speculate that the ballistic aspects of stemmed points were not highly sensitive to global climate dynamics and local geographic variation, but instead were optimal for the wide range of conditions encountered by people during the Late Palaeolithic in Korea.

We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
